Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
Done with first pass

In previous section, we presented a refunctionalized AAM with @nd@ operator
in extended continuation-passing style.
To obtain definitional abstract interpreters, we now transform it further to direct-style.

To do this at this point, we actually have multiple choices:
\begin{itemize}
  \item By monadifying the continuations into monads, we obtain the abstracting
    definitional interpreters in monadic style which are what \citeauthor{darais2017abstracting}
    described \cite{darais2017abstracting}.
  \item By representing the extended continuation-passing style with delimited controls,
    we derive a new form of abstract interpreters in direct-style.
  \item In fact, we may also just use the same caching algorithm but with side effect such as
    assignments and mutations to update the cache and then achieve the same
    definitional interpreter with pushdown control flows. In this case, nondeterminism can be
    well handled by for comprehension.
\end{itemize}

These coincidence should not be a surprise since there are extraordinarily rich literatures
show the correspondence between monads and continuation-passing style as well as delimited
controls since their very beginning era. \todo{cite}

In this section, we present the version that using delimited controls.

@@

As we identified before,

@@

May be insert to section refunc.

@@

For deterministic languages, a reduction semantics is a structural operational
semantics in continuation style, where the reduction context is a
defunctionalized continuation.
Unfortunately, the abstract abstract machines is nondeterministic, where
the nondeterminism is handled by a work list.
We can turn this explicit nondeterminism into the reduction context of
AAM.

\textbf{Pushdown Control flow Analysis.}
In recent years, there are significant efforts \cite{vardoulakis2010cfa2, earl2012introspective,
gilray2016pushdown, johnson2015abstracting} to achieve precise call/return
match based on small-step abstracting abstract machines.

CFA2 is the first solution that solves the return-flows problem\cite{vardoulakis2010cfa2}.
But CFA2 has several drawbacks: it works on only continuation-passing style programs,
and not supports polyvariant analysis, in addition with an exponential time
complexity.

@@

\citeauthor{gilray2016pushdown} proposed a polyvariant continuation-addresses allocator
for small-step AAM to achieve pushdown analysis \cite{gilray2016pushdown}.
The method is both simple to implement and computational inexpensive.
Based on the AAM we presented in Section \ref{aam}, the only change in the code is
not only keeping track of the entry-point expression of callee, but also holding the target environment
when allocating a continuation addresses.
The following code shows the implementation:

@@

The target environment is used for successor states and extended from closure's
environment (namely @val new_env = env_c + (v -> baddr)@).
All the other part of the code would not be changed.
Notably, this change only causes constant-factor increase in time complexity of
the analysis if the store is widened.

\textbf{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order
  programming languages} \cite{reynolds1972definitional} shows that in
definitional interpreters the defined language can inherit properties from the defining language.
With this insight, \citeauthor{darais2017abstracting} construct abstracting
definitional interpreters that automatically inherits the pushdown control-flow
property from its defining language
because the defined language just uses the stack the meta-language \cite{darais2017abstracting}.
\citeauthor{darais2017abstracting}'s abstracting definitional interpreters work
on direct-style programs, and is written in monadic style. One of the advantages
of monadic abstract interpreters is modular and composable, therefore deploying
different sensitivities or features is just applying a different monad.
Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979} show a monadic
abstract interpreter for small-step semantics.

This paper is inspired greatly from \citeauthor{darais2017abstracting}'s work.
Instead of just showing the final form of abstract definitional interpreters,
We reveal that refunctionalization plays an important role for inheriting
stack from the meta-language.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written in both
monadic- and direct-style.
The difference is ours work for A-Normal Form $\lambda$-calculus instead of plain
$\lambda$-calculus, our work also can be easily adopted to plain
$\lambda$-calculus.

>>>>

In the previous section, we presented a refunctionalized AAM utilizing the @nd@ operator
in extended continuation-passing style.
To obtain definitional abstract interpreters, we now transform it further to direct-style.

We have multiple choices of how to proceed:
\begin{itemize}
  \item By monadifying the continuations, we obtain the abstracting
    definitional interpreters in monadic style which are what \citeauthor{darais2017abstracting}
    describe \cite{darais2017abstracting}.
  \item By representing the extended continuation-passing style with delimited controls,
    we derive a new form of abstract interpreters in direct-style.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same
    definitional interpreter with pushdown control flows. In this case, nondeterminism can be
    easily handled for comprehension.
\end{itemize}

These coincidences should not be a surprise since existing literature is extraordinarily rich
in showing the correspondence between monads and continuation-passing style as well as
delimited controls since their very beginning era. \todo{cite}

In this section, we present the version that uses delimited controls.

@@

As identified before, \note{JD: compelling...}

@@

Maybe insert to section refunc.

@@

For deterministic languages, a reduction semantics is a structural operational
semantics in continuation style, where the reduction context is a
defunctionalized continuation.
Unfortunately, abstract abstract machines are nondeterministic, where
the nondeterminism is handled by a worklist.
We can turn this explicit nondeterminism into the reduction context of
AAM.

\textbf{Pushdown Control Flow Analysis.}
In recent years, there are significant efforts \cite{vardoulakis2010cfa2, earl2012introspective,
gilray2016pushdown, johnson2015abstracting} to achieve precise call/return
match based on small-step abstracting abstract machines.

CFA2 is the first solution that solves the return-flows problem\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs,
and does not supports polyvariant analysis, in addition to an exponential time
complexity.

@@

\citeauthor{gilray2016pushdown} proposed a polyvariant continuation-addresses allocator
for small-step AAM to achieve pushdown analysis \cite{gilray2016pushdown}.
This method is both simple to implement and computationally inexpensive.
Based on the AAM we presented in Section \ref{aam}, the only changes in the code required is
not only keeping track of the entry-point expression of callee, but also holding the target
environment when allocating a continuation addresses.
The following code shows the implementation:

@@

The target environment is used for successor states and extended from the closure's
environment (namely @val new_env = env_c + (v -> baddr)@).
No other pieces of code would need to be modified.
Notably, this change only causes constant-factor increase in time complexity of
the analysis if the store is widened.

\textbf{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order
  programming languages} \cite{reynolds1972definitional} shows that in
definitional interpreters, the defined language can inherit properties from the defining language.
With this insight, \citeauthor{darais2017abstracting} construct abstracting
definitional interpreters which automatically inherit the pushdown control-flow
property from its defining language
because the defined language simply uses the stack the
meta-language~\cite{darais2017abstracting}. \citeauthor{darais2017abstracting}'s abstracting
definitional interpreters work on direct-style programs, and is written in monadic style. One
of the advantages of monadic abstract interpreters is modular and composable, therefore
deploying different sensitivities or features is just applying a different monad.
Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979} show a monadic
abstract interpreter for small-step semantics.

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
Instead of simply showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.

The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written in both
monadic- and direct-style.
The difference is that ours works for A-Normal Form $\lambda$-calculus, instead of plain
$\lambda$-calculus, although our work can also be easily adopted to handle a plain
$\lambda$-calculus.
